---
title: "Final Project"
author: "Cierra Wickliff"
date: "2023-12-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width="95%")
```

# Load and subset 2020 Data

```{r}
## Use this chunk to load the data with read.csv()
worldbank <- read.csv("wdidata20.csv", stringsAsFactors = T)
worldbank2020 <- subset(worldbank, year==2019 & !is.na(SP.DYN.LE00.IN) & !is.na(income))
nrow(worldbank2020)
```


There are 197 countries that have life expectancy data from the year 2019.

# Number in Each Income Group

```{r}
## table
table(worldbank2020$income)
```

There are 65 High income, 30 low income, 47 lower middle income, and 55 upper middle income countries.


# Histogram of Life Expectancies

```{r}
## histogram
mean(worldbank2020$SP.DYN.LE00.IN)
median(worldbank2020$SP.DYN.LE00.IN)
hist(x=worldbank2020$SP.DYN.LE00.IN, 
     breaks=50,
     main = "Life Expectancy in Years")

```

The mean life expectancy is 72.93 years. The median is 74.24 years.
  

# Calculating the average life expectancy across income groups

```{r}
## tapply and barplot
incomeLife <- tapply(worldbank2020$SP.DYN.LE00.IN, worldbank2020$income, mean)
incomeLife
barplot(incomeLife,
        col=c("forestgreen", "red4", "orange2", "gold1"),
        border=NA,
        main="Life Expectancy by Income Group",
        ylab="Life Expectancy in Years")
```
Low income countries have a life expectancy of 62.82 years. Lower middle income countries have a life expectancy of 68.30 years. Upper middle income countries have a life expectancy of 73.91 years. High income countries have a life expectancy of 80.12 years.


# Quantify Income Groups

```{r}
## quantify data
worldbank2020$incomeNum <- NA
worldbank2020$incomeNum[worldbank2020$income=="Low income"] <- 1
worldbank2020$incomeNum[worldbank2020$income=="Lower middle income"] <- 2
worldbank2020$incomeNum[worldbank2020$income=="Upper middle income"] <- 3
worldbank2020$incomeNum[worldbank2020$income=="High income"] <- 4
```

I quantify the income groups from
  Low income: 1
  Lower middle income: 2
  Upper middle income: 3
  High income: 4


# Quantify Income Groups

```{r}
plot(x=worldbank2020$incomeNum, y=worldbank2020$SP.DYN.LE00.IN,
     main="Life Expectancy by Income Group",
     ylab="Income Group",
     xlab="Life Expectancy in Years")
```







